Reviving the Lost Art of Combinator Graph Reduction

### 2023-01-23 experiments/superg/doc/paper/paper.tex
Write comparison of TIGRE-style engine.

@@

\subsection{Panicking inside combinators}
Rust has the concept of \emph{panic}: an exceptional situation that indicates a bug in the program, for example, an index out-of-bounds error.
If a panic occurs, the program collects a stack trace and terminates with an error code.
A panic inside of a combinator implementation may cause the program to unwind into the node graph, which is undesirable because it is undefined behaviour\footnote{\url{https://doc.rust-lang.org/nomicon/ffi.html\#ffi-and-unwinding}}.
To prevent this, all combinator implementations are wrapped in a call to \emph{catch\_unwind}\footnote{\url{https://doc.rust-lang.org/std/panic/fn.catch_unwind.html}}, so we catch panics before they leave Rust code, log them and abort the program.

\section{Performance evaluation}
\label{sec:eval}
Programs:
\begin{itemize}
    \item ackermann
    \item digits of e
    \item linfib
    \item lsort
    \item nfib
    \item primes
    \item queens
    \item tak
    \item towers of hanoi
    \item treesort
\end{itemize}

@@

>>>>

\usepackage{svg}

\ifCLASSOPTIONcompsoc
    \usepackage[caption=false, font=normalsize, labelfont=sf, textfont=sf]{subfig}
\else
\usepackage[caption=false, font=footnotesize]{subfig}
\fi

@@

% Removed because it was too slow
%\subsection{Panicking inside combinators}
%Rust has the concept of \emph{panic}: an exceptional situation that indicates a bug in the program, for example, an index out-of-bounds error.
%If a panic occurs, the program collects a stack trace and terminates with an error code.
%A panic inside of a combinator implementation may cause the program to unwind into the node graph, which is undesirable because it is undefined behaviour\footnote{\url{https://doc.rust-lang.org/nomicon/ffi.html\#ffi-and-unwinding}}.
%To prevent this, all combinator implementations are wrapped in a call to \emph{catch\_unwind}\footnote{\url{https://doc.rust-lang.org/std/panic/fn.catch_unwind.html}}, so we catch panics before they leave Rust code, log them and abort the program.

\section{Performance evaluation}
\label{sec:eval}
Our performance evaluation consists of three parts:
\begin{itemize}
    \item We compare our Miranda-style engine with the original Miranda, to demonstrate that our implementation is competitive.
    \item We compare our TIGRE-style engine to our Miranda-style engine, to see if the techniques described in~\cite{koopman_fresh_1989} are viable on the X86 architecture.
    \item We compare the traditional bracket abstraction and the different semantic translation algorithms from~\cite{kiselyov_lambda_2018} by compiling the same source programs with each translation algorithm and measuring the runtime on our Miranda-style engine, to see if the semantic translation algorithms produce compiled programs that are faster to execute.
\end{itemize}

\subsection{Comparison with original Miranda}
Figure~\ref{fig:turner-vs-miranda} shows the performance of our Miranda-style engine with the traditional bracket abstraction compared to the original Miranda~\cite{noauthor_open_2021}.
The test program used for the benchmark is the function given in figure~\ref{fig:fib}.

\begin{figure*}
    \includesvg[width=\textwidth]{turner_vs_miranda}
    \centering
    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
    \label{fig:turner-vs-miranda}
\end{figure*}

\begin{figure}
    \begin{lstlisting}
-- Miranda
fib n = n,                       if n < 2
      = fib(n - 1) + fib(n - 2), otherwise

-- Superg (LISP-like)
(defun fib (n)
    (if (< n 2) 
        n
        (+ (fib (- n 1)) (fib (- n 2)))))
    \end{lstlisting}
    \centering
    \caption{
        Recursive function to compute the $n$-th element of the Fibonacci sequence.
    }
    \label{fig:fib}
\end{figure}

We observe that our Miranda-style engine is more than twice as fast as the original Miranda.
While we have not attempted an in-depth evaluation of the performance differences between these two systems, we can point to two likely reasons that our implementation performs better on this benchmark:
\begin{itemize}
    \item Miranda supports (exclusively) arbitrary precision integers.
          They are implemented by the high-performance GNU MP~\cite{gnump} library, but are still slower than the native 32-bit machine integers that our engine uses.
    \item As mentioned in~\ref{sec:miranda}, the original Miranda must be compiled with optimizations disabled.
          Without these optimizations, the compiler is likely to generate a slower implementation of key functions.
\end{itemize}

As the points above indicate, the comparison of the two systems is not fair.
Therefore, we do not claim that our implementation is strictly better than the original, but rather that it is a reasonably fast reference implementation of a combinator graph reduction engine.
The other benchmarks in this section all use the Miranda-style engine with bracket abstraction as a reference point for performance, based on which we can judge if the proposed changes are beneficial.

\subsection{TIGRE-style Engine}
To benchmark the TIGRE-style engine we use the same Fibonacci program and bracket abstraction algorithm, and plot the execution time against our Miranda-style engine.
Both engines are configured to use the same compiler, so they will execute identical combinator expressions.
The results can be seen in figure~\ref{fig:turner-vs-tigre}.

\begin{figure*}
    \includesvg[width=\textwidth]{turner_vs_tigre}
    \centering
    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
    \label{fig:turner-vs-tigre}
\end{figure*}

Contrary to our expectations, the TIGRE-style engine is significantly slower than the reference Miranda-style engine.
Based on the flame graphs~\cite{gregg_flame_2016} of the two engines, available in Appendix~\ref{app:flame-graph}, we can conclude that the S combinator is significantly slower on this engine.

We used Valgrind's cachegrind~\cite{NethercotePhd2004} tool to inspect the \texttt{make\_s} function and found that its execution time is dominated by data write operation to update the graph.
While the TIGRE-style engine executes fewer instructions (22M vs. 79M) and performs fewer data accesses (13M vs. 49M), it has a much higher cache miss rate for writes (3\% vs. 0.5\%).

We hypothesize that updates to nodes in the graph are slow in the TIGRE-style engine because those same nodes have been loaded into the instruction cache, and therefore trigger an expensive cache invalidation to maintain coherency between the instruction and data caches.
The node to be updated is part of the current spine, so its \texttt{CALL} instruction will have been executed before the update, loading it into the instruction cache.
When a write occurs to this node, the CPU must conservatively invalidate or update that cached value, or it would risk executing a stale instruction if the code were to jump back to that address.
Writes to addresses in the instruction cache are frequent in the TIGRE-style engine, but in most software, they are very rare.
On some processor architectures, like ARM, programs must manually clear the instruction cache if they issue a write to a location that may be cached~\cite{bramley_caches_2013}.

It appears that the TIGRE approach to implementing combinator graph reduction is a poor fit for modern architectures due to the split between instruction and data caches.
Other innovations from TIGRE, such as \texttt{LIT} combinators and removing tags, may still be beneficial.
In particular, it may be possible to get the benefits of TIGRE without cache coherency issues by storing jump addresses rather than call instructions in nodes.
Rather than jump directly to a node and execute it, the runtime could load the jump address and push the argument pointer to the stack normally, then jump to the address instead.
The implementation of such an engine is left as future work.

\subsection{Kiselyov's algorithms}

\textbf{TEXT BELOW ARE NOTES!}
Programs:
\begin{itemize}
    \item ackermann
    \item digits of e
    \item linfib
    \item lsort
    \item nfib
    \item primes
    \item queens
    \item tak
    \item towers of hanoi
    \item treesort
\end{itemize}

@@

\appendix

\subsection{Flame graphs}
\label{app:flame-graph}
\begin{figure*}
    \includesvg[inkscapeformat=png, width=\textwidth]{flamegraph_turner}
    \centering
    \caption{Flame graph for $fib(20)$ on Miranda-style engine.}
    \label{fig:flamegraph-turner}
\end{figure*}

\begin{figure*}
    \includesvg[inkscapeformat=png, width=\textwidth]{flamegraph_tigre}
    \centering
    \caption{Flame graph for $fib(20)$ on TIGRE-style engine.}
    \label{fig:flamegraph-tigre}
\end{figure*}
